y---
title: "R Coding Exercise"
---

Placeholder file for the future R coding exercise.

# Install all packages needed

Install and library all packages needed in this section.

```{r}
install.packages("dslabs") #if we have not install this packages, we need to indtall it. This packege is for the data set 

install.packages("tidyverse")
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
library(here)
install.packages("patchwork")  # This package is to redefine "/" operator for plot arrangement
library(patchwork)
```

# Loading Dataset

In this projec, we use data from dslabs.

```{r}
library(dslabs)

help("gapminder") # this is to look at the description about the data
```

let's look at the overview of the data structure

```{r}
str(gapminder)

```

Let's take a look a summary of the data

```{r}
summary(gapminder)
```

Now, we want to check the type of object gapminder

```{r}
class(gapminder)
```

We can see that the object gapminder is a data frame

# Processing data

To start with, I want to create a new data set/ object called africadata, which I create by selecting from African countries only.

```{r}
africadata <- filter(gapminder, continent== "Africa") # filter() function is used to filter specific observation only. 
```

Now, I am going to check the new dataset using str() function

```{r}
str(africadata)
```

Now, we can see that the number of variables is exactly same with the number of variable in gapminder dataset, but the number of observation only 2907 instead of 10545 as in gapminder. It is because in previous step, we try to select countries from African continent only.

Now let see the summary of the data using summary() function

```{r}
summary(africadata)
```

From the summary, we can see a simple summary of the data including number minimum, first Qu, median, mean, 3rd qu, and maximum value of numeric variables.

From africadata, I want to create two new objects called infant and pop (stand for population). Object infant contains infant_mortatility and life_expectancy variables, and pop contains population and life_expectancy variables.

```{r}
infant <- select(africadata, infant_mortality, life_expectancy)
str(infant)

summary(infant)
```

```{r}
pop <- select(africadata, population, life_expectancy)
str(pop)
summary(pop)
```

# Plotting

In this section, I am going to create some plots.

The first plot is a plot of life expectancy as a function of infant mortality in African countries.

```{r}
plot1 <- ggplot(data = infant, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point(col = "purple") + # to create plot points and set color to purple  
  geom_smooth(method = "loess", se = FALSE) + # to create line without confidence interval
  xlim(c(0, 300)) + # to set x-axis range
  ylim(c(0, 100)) + # to set y-axis range
  labs(
    subtitle = "(Figure 1. Life Expectancy as a Function of Infant Mortality in Africa)", 
    y = "Life Expectancy (Years)",
    x = "Infant Mortality (per 1000)",
    caption = "Source: Gapminder Data"
  ) + # to create the name of each axis
  theme(
    plot.background = element_rect(color = "black", size = 1), # Border around the entire plot (including title)
    plot.subtitle = element_text(hjust = 0.5, vjust = -1, size = 12, color = "gray50", face = "bold.italic"), # to adjust the subtitle position, color and size (note, set vjust=185 for printed pictures, for website, set at 100 )
    plot.caption = element_text(hjust = 1, vjust = -13, size = 10, color = "gray40"), # to adjust the subtitle position, color and size, hjust=horizontal and vjust= vertical 
    plot.margin = margin(10, 10, 45, 10) # (top, right, bottom, left) Increase the bottom margin to create space for the subtitle below the x-axis
  )

# Print the plot
print(plot1)


figure_file = here("images","infant mortality vs life expectancy.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot1) # save the pictures created 
```

From figure 1, it can be seen that there is a negative relationship between infant mortality rate and life expectancy, the higher the infant mortality the lower life expectancy of a country.

Now, I want to see the relationship between population and life expectancy. In this plot, we will set population on log scale.

```{r}
plot2 <- ggplot(data = pop, aes(x= population, y= life_expectancy))+
  geom_point(color= "blue")+
  scale_x_log10() + # this is to set the x-axis in log scale 
  labs(
    subtitle = "(Figure 2. Life Expectancy as a Function of Population in Africa)", 
    y = "Life Expectancy (Years)",
    x = "Country Population (log scale)",
    caption = "Source: Gapminder Data"
  ) + # to create the name of each axis
  theme(
    plot.background = element_rect(color = "black", size = 1), # Border around the entire plot (including title)
    plot.subtitle = element_text(hjust = 0.5, vjust = -1, size = 12, color = "gray50", face = "bold.italic"), # to adjust the subtitle position, color and size (note, set vjust=185 for printed pictures, for website, set at 100 )
    plot.caption = element_text(hjust = 1, vjust = -13, size = 10, color = "gray40"), # to adjust the subtitle position, color and size, hjust=horizontal and vjust= vertical 
    plot.margin = margin(10, 10, 45, 10) # (top, right, bottom, left) Increase the bottom margin to create space for the subtitle below the x-axis
  )
print(plot2)


figure_file = here("images","log_population vs life expectancy.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot2) # save the pictures created 
```

From figure 2, overall there is a positive relationship between country population and life expectanc. However, the data look steak.

There are several factors that might affect this trend. First, the data is over 56 year period (1960 to 2016), where the number of population in most of the country increasing and health condition improving. To get better understanding about this figure, we need to include other variables such as GDP, fertility and other variables.

# More data processing

In this step, I am going to find out which year

```{r}
missing_years <- africadata %>%
  filter(is.na(infant_mortality)) %>% # Filter rows where infant_mortality is NA
  select(year) %>%                    # Select the 'year' column
  distinct()                          # Get unique years

# Display missing years
print(missing_years)
```

There are 23 years with missing data in infant mortality

Now, I want to create a new object containing data of the year 2000 only.

```{r}
data_2000 <- africadata %>%
  filter(year==2000)

```

```{r}
str(data_2000)
summary(data_2000)
```

there are 51 country selected in 2000 without missing data for child mortatlity

```{r}
plot3 <- ggplot(data = data_2000, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point(aes(color = region)) + # Map 'region' to the color aesthetic
  geom_smooth(method = "loess", se = FALSE) + # Add a smoothing line
  labs(
    subtitle = "(Figure 3. Life Expectancy as a Function of Infant Mortality in Africa in 2000)", 
    y = "Life Expectancy (Years)",
    x = "Infant Mortality (per 1000)",
    caption = "Source: Gapminder Data"
  ) +
  theme(
    plot.background = element_rect(color = "black", size = 1), # Border around the entire plot
    plot.subtitle = element_text(hjust = 0, vjust= 0, size = 12, color = "gray50", face = "bold.italic"), # Adjust subtitle position
    plot.caption = element_text(hjust = 1, vjust= -13, size = 10, color = "gray40"), # Adjust caption position
    plot.margin = margin(10, 10, 45, 10) # Adjust margins (top, right, bottom, left)
  )

print(plot3)

figure_file = here("images","infant mortality vs life expectancy in 2000.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot3) # save the pictures created 
```

Let see the plot for life expectancy vs country population

```{r}
plot4 <- ggplot(data = data_2000, aes(x= population, y= life_expectancy))+
  geom_point(aes(color = region)) +
  scale_x_log10() + # this is to set the x-axis in log scale 
  labs(
    subtitle = "(Figure 4. Life Expectancy as a Function of Population in Africa in 2000)", 
    y = "Life Expectancy (Years)",
    x = "Country Population (log scale)",
    caption = "Source: Gapminder Data"
  ) + # to create the name of each axis
  theme(
    plot.background = element_rect(color = "black", size = 1), # Border around the entire plot (including title)
    plot.subtitle = element_text(hjust = 0.5, vjust = -1, size = 12, color = "gray50", face = "bold.italic"), # to adjust the subtitle position, color and size (note, set vjust=185 for printed pictures, for website, set at 100 )
    plot.caption = element_text(hjust = 1, vjust = -13, size = 10, color = "gray40"), # to adjust the subtitle position, color and size, hjust=horizontal and vjust= vertical 
    plot.margin = margin(10, 10, 45, 10) # (top, right, bottom, left) Increase the bottom margin to create space for the subtitle below the x-axis
  )
print(plot4)


figure_file = here("images","log_population vs life expectancy in 2000.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot4) # save the pictures created 
```

From figure 3 and figure 4, the points of the plots are more clear because the data is only for the year 2000. In these figures, I use different color based on the region, it can help visualize some figures. For example, we can see that overall, Northern Africa region generally had higher life expectancy compared to other regions, while eastern region had lower life expectancy.

# Fitting simple models

In this part, I want to perform simple linear model to look at the relationship between infant mortality and life expectancy, and country population on life expectancy.

```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, data = data_2000)
summary(fit1)
```

Based on the simple linear model, the intercept is 71.4, meaning that the average life expectancy is 71.4 years when the infant mortality is 0. It is estimated that every 1000 increase child mortality results in decrease life expectancy by 0.2 year Based on the alpha=0.05, it can be concluded that there is a statistically significant relationship between infant mortality rate and life expectancy with p\<.0001.

```{r}
fit2 <- lm(life_expectancy ~ population, data = data_2000)
summary(fit2)
```

The second model explain the relationship between life expectancy and country population. Based on the intercept, it is estimated that when country population is zero, the average life expectancy is 5.6 years, but in the reality, it does not make sense since none of country has 0 population. Based on the alpha= 5% there is no significant relationship between country population and life expectancy with p=0.616.


## Natalie's Addition to Muhammad's Exercise 3

This section is contributed to by Natalie Cann. 

I will first load packages needed for this exercise. 
```{r}
library(dslabs)
library(ggplot2)
library(tidyverse)
library(dplyr)
```
I will use the "help()" function on the "us_contagious_diseases" data from dslabs.
```{r}
help(us_contagious_diseases)
```
The help file told me that this data frame contains yearly counts for Hepatitis A, Measles, Mumps, Pertussis, Polio, Rubella, and Smallpox in the United States. 

Now, I will use str() on the "us_contagious_diseases" data frame. 
```{r}
str(us_contagious_diseases)
```
This informs me that the data set contains 6 variables (disease, state, year, weeks_reporting, count, and population) and 16,065 observations. Disease and state are factors, while the other four variables are numeric. 

Next, I will use the summary() on the "us_contagious_diseases" data frame. 
```{r}
summary(us_contagious_diseases)
```
The summary above shows the diseases and states. It also shows the minimum, 1st quartile, median, mean, 3rd quartile and maximum values for the numeric variables (year, weeks_reporting, count, and population). 

First, I will create a data frame containing only data for the state of Georgia. 
```{r}
GA_contagious_diseases <- filter(us_contagious_diseases, state == "Georgia") # I am using filter to obtain only data from GA
View(GA_contagious_diseases) # I am viewing the data frame to ensure it only contains data from GA
```
This worked, as only GA data is shown in the new data frame. 

Now, I will run str() and summary() on the GA_contagious_diseases data frame. 
```{r}
str(GA_contagious_diseases)
summary(GA_contagious_diseases)
```
This shows me that all the variables are the same as before, but they only contain data from GA. 

Now, I will create a data frame that contains only data from GA in 1950. I will use this data frame to create a plot of the number of cases of each disease in GA in 1950. 
```{r}
GA_contagious_diseases_1950 <- filter(GA_contagious_diseases, year == 1950) # I am using filter to obtain only data from GA in 1950

# I will run str() and summar() on the GA_contagious_diseases_1950 data frame to get a better look at the data
str(GA_contagious_diseases_1950) 
summary(GA_contagious_diseases_1950)
```
You can see that now, the only year appearing in the data set is 1950. Therefore, all the variables now reflect only data from 1950 in GA. 

Next, I will rename the "count" variable to "number_of_cases". 
```{r}
GA_contagious_diseases_1950 <- rename(GA_contagious_diseases_1950, number_of_cases = count) # I am renaming the "count" variable to "number_of_cases" via the rename() function (and then assigning it back to GA_contagious_diseases_1950)

colnames(GA_contagious_diseases_1950) # I am checking to see if the previous step was done properly
```
I see this was done correctly, as the variable that used to be called "count" is now called "number_of_cases". 

Now, I will create a bar graph to display the number of cases of each contagious disease reported in GA in 1950. 
```{r}
custom_colors_1950 <- c("Measles" = "#6fe51e", "Pertussis" = "#2ce1b0", "Polio" = "#2cb2e1", "Smallpox" = "#3a83e6") # I am creating a vector of colors that I will use to fill each disease's bar on the graph below

ggplot(GA_contagious_diseases_1950, aes (x = disease, y = number_of_cases, fill = disease)) + # Using ggplot on the GA_contagious_diseases_1950 data frame and setting x and y equal to diseases and number of cases (respectively) and setting fill equal to disease
  geom_bar(stat = "identity") + # Specifying the geom as geom_bar() to create a bar graph 
  labs(title = "Reported Number of Cases of Contagious Diseases \n in State of Georgia in 1950", x = "Disease", y = "Number of Cases") + # Renaming title and axes
  scale_fill_manual(values = custom_colors_1950) + # Setting the fill colors of the bars to the custom colors I created above 
  geom_text(aes(label = number_of_cases), vjust = -0.5) + # Adding text labels to the top of each bar to show the number of cases
  theme(legend.position = "bottom", plot.title = element_text(size = 18, face = "bold", hjust = 0.5), axis.title.x = element_text(size = 12, face = "bold"), axis.title.y = element_text(size = 12, face = "bold")) + # Putting legend at the bottom; increasing size, boldness, and center of title/axes
  scale_y_continuous(limits = c(0, 2500), breaks = seq(0, 2500, by = 500)) # Setting the y-axis limits and breaks to better see the text of number of cases at the top
```
You can see from the bar graph above that the number of Measles cases in GA in 1950 was the highest out of all the diseases shown here (with the number of cases being 2159). 

Now, I will create a data frame that contains Measles data from all states in 1950. 
```{r}
Measles_1950 <- filter(us_contagious_diseases, disease == "Measles", year == 1950) # I am using filter to obtain only Measles data from all states in 1950

# I will run str() and summary() on the Measles_1950 data frame to get a better look at the data
str(Measles_1950) 
summary(Measles_1950)
```
This shows data only from the Measles disease; it includes data from every state that reported measles data in 1950. 

Once again, I will rename the "count" variable to "number_of_cases". 
```{r}
Measles_1950 <- rename(Measles_1950, number_of_cases = count) # I am renaming the "count" variable to "number_of_cases" via the rename() function 

colnames(Measles_1950) # Making sure this was done correctly
```
As seen by the output, I can see that this variable is now called "number_of_cases". 

Now, I will create a bar graph to display the number of Measles cases reported in each state in 1950. 
```{r}
custom_colors_Measles_1950 <- c("#53d127", "#71f45d", "#2ce1b0", "#40dcd9", "#2cb2e1", "#3a83e6")

ggplot(Measles_1950, aes(x = state, y = number_of_cases, fill = number_of_cases)) + 
  geom_bar(stat = "identity") +  # Correct stat for using y data directly
  labs(title = "Reported Number of Measles Cases \n in the United States in 1950", 
       x = "State", 
       y = "Number of Cases") + # Rename title and axes
  geom_text(aes(label = number_of_cases), vjust = -0.5, size = 1.5) +  # Display labels above bars
  theme(legend.position = "none",  # Remove legend 
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"), # Increase size and boldness of title and axes
        axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate state names
  scale_fill_gradientn(colors = custom_colors_Measles_1950) # Set custom color scheme made
```
This bar graph shows the number of Measles cases reported in each state in 1950. As you can see, Michigan had the highest number of reported cases of Measles; on the other hand, Alaska had the lowest number of reported cases of Measles. 

Now, I will create a new variable within the Measles_1950 data frame that assess number of cases in each region of the country. 
```{r}
Measles_1950$region <- ifelse(Measles_1950$state %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts", "Rhode Island", "Connecticut", "New York", "New Jersey", "Pennsylvania", "Delaware", "Maryland", "District Of Columbia"), "Northeast", # contains states from the Northeast
                        ifelse(Measles_1950$state %in% c("Ohio", "Michigan", "Indiana", "Illinois", "Wisconsin", "Minnesota", "Iowa", "Missouri", "North Dakota", "South Dakota", "Nebraska", "Kansas"),  "Midwest", # contains states from the Midwest
                        ifelse(Measles_1950$state %in% c("Virginia", "West Virginia", "North Carolina", "South Carolina", "Georgia", "Florida", "Kentucky", "Tennessee", "Alabama", "Mississippi", "Arkansas", "Louisiana"), "South", # contains states from the South
                        ifelse(Measles_1950$state %in% c("Texas", "Oklahoma", "New Mexico", "Arizona"), "Southwest", # contains states from the Southwest
                        ifelse(Measles_1950$state %in% c("Alaska", "Hawaii"), "Non-Contiguous", # contains states not connected to the main US
                        "West"))))) # any other state will be categorized as the West

head(Measles_1950) # I am checking to ensure this was done properly
```
I will create a version of this data frame that contains the number of cases of Measles in each region of the US. 
```{r}
region_Measles_cases_1950 <- Measles_1950 %>% # Creating new data frame that groups data by region and then includes the total number of cases in each region
  group_by(region) %>%
  summarize(total_cases = sum(number_of_cases))
```
 
Now, we can create a bar graph to display the number of Measles cases reported in each region in 1950. 
```{r}
ggplot(region_Measles_cases_1950, aes(x = region, y = total_cases, fill = total_cases)) +  # Using ggplot on the region_Measles_cases_1950 data frame and setting x and y equal to region and total_cases (respectively) and setting fill equal to region
  geom_bar(stat = "identity") +  # Use 'identity' since we are providing y values directly
  labs(title = "Reported Measles Cases by Region in 1950", 
       x = "Region", 
       y = "Number of Cases") + # Rename title and axes
  geom_text(aes(label = total_cases), vjust = -0.5, size = 3) +  # Add labels above bars
  theme(legend.position = "right", plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"), # Increase size and boldness of title and axes
        axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate state names
  scale_y_continuous(limits = c(0, 130000), breaks = seq(0, 130000, by = 20000)) + # Setting the y-axis limits and breaks to better see the text of number of cases at the top
  scale_fill_gradientn(colors = custom_colors_Measles_1950) # Set custom color scheme made

```
As you can see, the Midwest had the highest number of reported Measles cases in 1950 (115836). The Non-Contiguous region had the lowest number of reported Measles cases in 1950 (0) - it is important to note that Alaska and Hawaii are the only states in this region and that Alaska did not report any cases. 

I will create a Scatterplot of the number of cases of Measles in each state in 1950. 
```{r}
Measles_1950$state <- factor(Measles_1950$state, 
                             levels = unique(Measles_1950$state[order(Measles_1950$region)]))
# I am reordering the states based on the region they are in (instead of alphabetical order) via levels(), unique(), and order()

ggplot(Measles_1950, aes(x = state, y = number_of_cases, color = region)) + 
  geom_point(size = 3) +  # Specify geom as geom_point to make a scatterplot
  labs(title = "Reported Measles Cases in Each State in 1950", 
       x = "State", 
       y = "Number of Cases") +  # Rename title and axes
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate state names 
        legend.position = "bottom",  # Position legend at the bottom
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold")) +  # Increase size and boldness of title and axes
  scale_color_manual(values = c("Northeast" = "#53d127", "Midwest" = "#71f45d", "South" = "#2ce1b0", 
                               "Southwest" = "#47e2e7", "Non-Contiguous" = "#2cb2e1", "West" = "#3a83e6"))  # Assign specific colors to each region
```
As you can see from the scatterplot above, the Midwest and the Northeast have the greatest amount of variation in the number of reported Measles cases in 1950. 

I will now use the lm() function to fit a linear model with number_of_cases as the outcome and region of the United States as the predictor. Then, I will apply the summary() function to view the results. 
```{r}
fit_1950 <- lm(number_of_cases ~ region, data = Measles_1950) # Running a linear fit via lm()
summary(fit_1950) # Viewing results of linear model via summary()
```
The P-value of this model is greater than 0.05, which means we must fail to reject the null hypothesis (that there is no relationship between the number of Measles cases and the region of the US). This means that the region of the US does not have a significant impact on the number of Measles cases reported in 1950. 

I will run an ANOVA (Analysis of Variance) test to asses if there are significant differences in the means of the number of Measles cases reported in each region in the US in 1950.
```{r}
anova_result <- aov(number_of_cases ~ region, data = Measles_1950) # Running ANOVA test with aov() function
summary(anova_result) # Viewing a summary of results
```
The F-value of the ANOVA table is 1.386, and the P-value is 0.248. Since 0.248 > 0.05, we must fail to reject the null hypothesis, which is stated previously. Therefore, this result is consistent with the linear fit model results above; region of the US does not appear to have a significant impact on the number of Measles cases in 1950. 

I will now put the summary into a table. 
```{r}
anova_df <- as.data.frame(summary(anova_result)[[1]]) # Putting the summary of the ANOVA test into a data frame 
print(anova_df) # To Printing the  data frame
```
Here is the printed ANOVA results. 