---
title: "Data Exercise" 
output: html
editor: visual
---

# About the dataset

The data is obtained from The University of Michigan Health and Retirement Study (HRS). HRS is a longitudinal panel study that surveys a representative sample of approximately 20,000 people in America, supported by the National Institute on Aging (NIA U01AG009740) and the Social Security Administration.

Through its unique and in-depth interviews, the HRS provides an invaluable and growing body of multidisciplinary data that researchers can use to address important questions about the challenges and opportunities of aging.

Please visit: https://hrsdata.isr.umich.edu/data-products/public-survey-data?\_gl=1*8ofhfg*\_ga*MTc5MDM4ODE3NS4xNzM4MzU5MjU5*\_ga_FF28MW3MW2\*MTczODM1OTI1OS4xLjEuMTczODM1OTI3NC4wLjAuMA..

The data is in STATA format.

I plan to conduct survival analysis for retired people who had high Cholesterol in 2014 until they Heart Attack.

# Install all packages needed

Install and library all packages needed in this section.

```{r}


install.packages("tidyverse")
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
library(here)
install.packages("patchwork")  # This package is to redefine "/" operator for plot arrangement
library(patchwork)
install.packages("writexl")
library(writexl)
library(haven)

```

# Loading Dataset

This dataset is longitudinal survey. In this analisys, I take three years, 2014, 2016,and 2018. I will read data per year then combine them.

In this part, I will load the dataset using here() function.

Reading data for 2014 and 2016 needs several steps, because the data is in STATA format, and also we need to match with dictionary. \### Data for 2014

```{r}
## set path  dictionary 
dict.file2014 <- here::here("data-exercise","data", "H14C_R.dct")

file.exists(here::here("data-exercise", "data", "H14C_R.dct")) # If we want to check the working path 

# Read the dictionary file 
df.dict2014 <- read.table(dict.file2014, skip = 1, fill = TRUE, stringsAsFactors = FALSE)

# Set column names for dictionary dataframe 
colnames(df.dict2014) <- c("col.num","col.type","col.name","col.width","col.lbl")

# Remove last row which only contains a closing }
df.dict2014 <- df.dict2014[-nrow(df.dict2014),]


# Remove first row 
df.dict2014 <- df.dict2014[-1,]

# Extract numeric value from column width field
df.dict2014$col.width <- as.integer(sapply(df.dict2014$col.width, gsub, pattern = "[^0-9\\.]", replacement = ""))


# Convert column types to format to be used with read_fwf function
df.dict2014$col.type <- sapply(df.dict2014$col.type, function(x) ifelse(x %in% c("int","byte","long"), "i", ifelse(x == "float", "n", ifelse(x == "double", "d", "c"))))



##data file 
data.file2014 <- here::here( "data-exercise","data", "H14C_R.da") 

# Read the data file into a dataframe
df2014 <- read_fwf(
    file = data.file2014,
    fwf_widths(widths = df.dict2014$col.width, col_names = df.dict2014$col.name),
    col_types = paste(df.dict2014$col.type, collapse = ""))

# Add column labels to headers
#attributes(df2014)$variable.labels <- df.dict2014$col.lbl

head(df2014)
```

### Data for 2016

```{r}
## set path  dictionary 
dict.file2016 <- here::here("data-exercise","data", "H16C_R.dct")

file.exists(here::here("data-exercise", "data", "H16C_R.dct")) # If we want to check the working path 

# Read the dictionary file 
df.dict2016 <- read.table(dict.file2016, skip = 1, fill = TRUE, stringsAsFactors = FALSE)

# Set column names for dictionary dataframe 
colnames(df.dict2016) <- c("col.num","col.type","col.name","col.width","col.lbl")

# Remove last row which only contains a closing }
df.dict2016 <- df.dict2016[-nrow(df.dict2016),]


# Remove first row 
df.dict2016 <- df.dict2016[-1,]

# Extract numeric value from column width field
df.dict2016$col.width <- as.integer(sapply(df.dict2016$col.width, gsub, pattern = "[^0-9\\.]", replacement = ""))


# Convert column types to format to be used with read_fwf function
df.dict2016$col.type <- sapply(df.dict2016$col.type, function(x) ifelse(x %in% c("int","byte","long"), "i", ifelse(x == "float", "n", ifelse(x == "double", "d", "c"))))



##data file 
data.file2016 <- here::here( "data-exercise","data", "H14C_R.da") 

# Read the data file into a dataframe
df2016 <- read_fwf(
    file = data.file2016,
    fwf_widths(widths = df.dict2016$col.width, col_names = df.dict2016$col.name),
    col_types = paste(df.dict2016$col.type, collapse = ""))

# Add column labels to headers
#attributes(df2016)$variable.labels <- df.dict2016$col.lbl 
head(df2016)
```

### Data for 2018

Data for 2018 is simple to read because it has R data format. It does not need many steps needed as data for 2014 and 2016.

```{r}
df2018 <- here::here("data-exercise","data", "2018_h18c_r.dta")
df2018 <- read_dta(df2018)
head(df2018)
```

### Data for 2020

```{r}
df2020 <- here::here("data-exercise","data", "2020_H20C_R.dta")
df2020 <- read_dta(df2020)
head(df2020)
```

### Data for 2022

The data is in STATA format

```{r}
## set path  dictionary 
dict.file2022 <- here::here("data-exercise","data", "H22C_R.dct")

file.exists(here::here("data-exercise", "data", "H22C_R.dct")) # If we want to check the working path 

# Read the dictionary file 
df.dict2022 <- read.table(dict.file2022, skip = 1, fill = TRUE, stringsAsFactors = FALSE)

# Set column names for dictionary dataframe 
colnames(df.dict2022) <- c("col.num","col.type","col.name","col.width","col.lbl")

# Remove last row which only contains a closing }
df.dict2022 <- df.dict2022[-nrow(df.dict2022),]


# Remove first row 
df.dict2022 <- df.dict2022[-1,]

# Extract numeric value from column width field
df.dict2022$col.width <- as.integer(sapply(df.dict2022$col.width, gsub, pattern = "[^0-9\\.]", replacement = ""))


# Convert column types to format to be used with read_fwf function
df.dict2022$col.type <- sapply(df.dict2022$col.type, function(x) ifelse(x %in% c("int","byte","long"), "i", ifelse(x == "float", "n", ifelse(x == "double", "d", "c"))))



##data file 
data.file2022 <- here::here( "data-exercise","data", "H22C_R.da") 

# Read the data file into a dataframe
df2022 <- read_fwf(
    file = data.file2022,
    fwf_widths(widths = df.dict2022$col.width, col_names = df.dict2022$col.name),
    col_types = paste(df.dict2022$col.type, collapse = ""))

# Add column labels to headers
#attributes(df2022)$variable.labels <- df.dict2022$col.lbl #
head(df2022)
```

# Data Cleaning

## Select variables needed for each year

Variables that needed in this study include Household Identification number, Respondent Person Identification number, High Cholesterol Status, Years first had heart attack, Month first had heart attack, Ever had heart attack.

```{r}
#data2014
da2014 <- df2014 %>% 
  select("HHID", "PN", "OC283", "OC040", "OC258", "OC259") %>%
  rename(hhid = HHID, pn = PN, chol= OC283, ha = OC040, year = OC258, month= OC259)


da2014 <- lapply(da2014, function(x) { attributes(x) <- NULL; x }) #delete the column lables from dataset, because their position is incorrect after deleting some variables. 
da2014 <- as.data.frame(da2014)  # Convert back to a dataframe

#data2016
da2016 <- df2016 %>% 
  select("HHID", "PN", "PC283", "PC040", "PC258", "PC259") %>%
  rename(hhid = HHID, pn = PN, chol= PC283, ha = PC040, year = PC258, month= PC259)


da2016 <- lapply(da2016, function(x) { attributes(x) <- NULL; x }) #delete the column lables from dataset, because their position is incorrect after deleting some variables. 
da2016 <- as.data.frame(da2016)  # Convert back to a dataframe

#data2018 
da2018 <- df2018 %>%
  select("hhid", "pn", "QC283", "QC040", "QC258", "QC259") %>%
  rename( chol = QC283, ha= QC040, year = QC258, month = QC259) # I rename the column names because in this year, they use different name. 

da2018 <- lapply(da2018, function(x) { attributes(x) <- NULL; x }) #delete the column lables from dataset, because their position is incorrect after deleting some variables. 
da2018 <- as.data.frame(da2018)  # Convert back to a dataframe

da2020 <- df2020 %>%
  select("HHID", "PN", "RC283", "RC040", "RC258", "RC259") %>%
  rename( hhid = HHID, pn = PN, chol = RC283, ha= RC040, year = RC258, month = RC259) # I rename the column names because in this year, they use different name. 

da2022 <- df2022 %>%
  select("HHID", "PN", "SC283", "SC040", "SC258", "SC259") %>%
  rename( hhid = HHID, pn = PN, chol = SC283, ha= SC040, year = SC258, month = SC259) # I rename the column names because in this year, they use different name. 
```

## Combine dataset

```{r}
# Perform a full join step-by-step
merged_data <- da2014 %>%
  full_join(da2016, by = c("hhid", "pn"), suffix = c("_2014", "_2016")) %>%
  full_join(da2018, by = c("hhid", "pn"), suffix = c("", "_2018")) %>%
  full_join(da2020, by = c("hhid", "pn"), suffix = c("", "_2020")) %>%
  full_join(da2022, by = c("hhid", "pn"), suffix = c("", "_2022")) 

# View result
head(merged_data)
```

I realize that we only need cholesterol status in 2014, therefore, I want to delete cholesterol in 2016 and 2018

```{r}
df <- merged_data[, !(colnames(merged_data) %in% c("chol_2016", "chol", "chol_2020", "chol_2022"))] %>% # deleting cholesterol from 2016 and 2018 
  rename(ha_2018 = ha, year_2018 =year, month_2018 = month) # I rename variables from 2018 
```

# Exploratory data

I want to create pie chat to show persons having high cholesterol in 2014.

```{r}
table_1 <- table(df$chol_2014)

chol_df <- as.data.frame(table_1) %>% 
  select(1,2)

colnames(chol_df) <- c("Chol_stat", "Count") # rename the column 

chol_df  <- chol_df  %>%
  filter(!Chol_stat %in% c(8, 9)) %>% 
  mutate(Chol_category = case_when(
    Chol_stat == 1 ~ "Normal",  # If Chol_stat is 1, it's "Normal"
    Chol_stat == 5 ~ "High",    # If Chol_stat is 5, it's "High"
    TRUE ~ "Other"              # For any other value, label as "Other" (optional)
  )) %>%
  select(Chol_category, Count) %>%
  rename(Chol_stat = Chol_category)
  

plot1 <- ggplot(chol_df, aes(x = "", y = Count, fill = Chol_stat)) + 
  geom_bar(stat = "identity", width = 1) +  # Create bar chart
  coord_polar(theta = "y") +  # Convert to pie chart
  geom_text(aes(label = paste0(round(Count / sum(Count) * 100, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white") +  # Add percentage labels
  labs(title = "Cholesterol status in 2014") +
  theme_void()


figure_file = here("data-exercise","result","figures","chol_status.png")
ggsave(filename = figure_file, plot=plot1) 

print(plot1)

```

I want to create plot number of person having Heart Attack in each year

```{r}
table_2 <- table(df$ha_2014)
table_3 <- table(df$ha_2016)
table_4 <- table(df$ha_2018)
table_5 <- table(df$ha_2020)
table_6 <- table(df$ha_2022)


# Create the combined table
table_ha <- data.frame(
  status = c("1"),  # Heart attack status
  `2014` = table_2[c("1")],  # Heart attack status counts for 2014
  `2016` = table_3[c("1")],  # Heart attack status counts for 2016
  `2018` = table_4[c("1")],  # Heart attack status counts for 2018
  `2020` = table_5[c("1")],  # Heart attack status counts for 2020
  `2022` = table_6[c("1")]   # Heart attack status counts for 2022
) %>%
  rename(ha_2014 =X2014, ha_2016=X2016, ha_2018=X2018, ha_2020 =X2020, ha_2022 =X2022) %>%
  mutate(ha_case = ifelse(status==1, "heart attack", "No heart attack")) %>%
  select(ha_case, everything()) %>%  # Move ha_case to the first column
  select(-status) %>%  # Remove the status column
  select(-ha_case) # I delete it because it prevent to create long tabel 

# Create long table 
table_ha <- table_ha %>%
  pivot_longer(cols = starts_with("ha_"),  # Reshape the ha_ columns into long format
               names_to = "year",  # Column for years
               values_to = "count") %>%
  select(year, count)  

# I want to change the value into numeric year
table_ha <- table_ha %>% 
  mutate(years = case_when(
    year == "ha_2014" ~ "2014",
    year == "ha_2016" ~ "2016",
    year == "ha_2018" ~ "2018", 
    year == "ha_2020" ~ "2020",
    year == "ha_2022" ~ "2022"
  )) %>%
  select(years, everything())

table_ha <- table_ha %>% 
  select(-year)
head(table_ha)
```

After creating long table, I want to create plot to show the number of heart attack cases each observation year.

```{r}
plot2 <- ggplot(table_ha, aes(x = years, y = count)) +
  geom_line(size = 1, color = "blue") +
  geom_point(size = 3, color = "red") +
  labs(title = "Heart Attack Cases Over Time Among Retired People in the USA ", 
       x = "Years", 
       y = "Number of Heart Attack Cases ",
       caption= "Sourse: Health and Retirement Study") +
  theme_minimal() + 
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = "bold.italic")) + # to adjust the title position, size, and color
   theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1)) # to create frame of the plot 


figure_file = here("data-exercise","result","figures","heart attack cases.png")
ggsave(filename = figure_file, plot=plot2) 

print(plot2) 
```

Now, I want to perform inclusion and inclusion criteria. I will include person who had high cholesterol level in 2014, but they did not have heart attack (disease free) in 2014.

```{r}
# Step 1: Filter the dataset

df_filtered <- df %>%
  filter(chol_2014 == 1, # High cholesterol in 2014
         ha_2014 == 5) # No heart attack in 2014
data <- df_filtered

```
