---
title: "CDC Data Exercise" 
output: html
editor: visual
---

# About the dataset

The Botswana Combination Prevention Project (BCPP) was a collaborative research effort led by the Botswana Ministry of Health (MOH), the Harvard School of Public Health/Botswana Harvard AIDS Institute Partnership (BHP), and the U.S. Centers for Disease Control and Prevention (CDC). This community-based randomized trial aimed to assess the impact of various HIV prevention strategies on reducing HIV incidence across 15 intervention and 15 control communities. The intervention communities received comprehensive HIV testing, linkage to care, and universal treatment services, guided by the UNAIDS 90-90-90 targets: ensuring 90% of individuals with HIV are aware of their status, 90% of those diagnosed are on antiretroviral therapy (ART), and 90% of those on ART achieve viral suppression.

The BCPP study was structured around two interrelated protocols: the Evaluation Protocol and the Intervention Protocol. The Evaluation Protocol assessed the primary outcome—HIV incidence—along with key secondary outcomes, focusing on data collected from the Baseline Household Survey, the HIV Incidence Cohort, and the End of Study Survey. Meanwhile, the Intervention Protocol involved the implementation of a combination prevention (CP) package in combination prevention communities (CPCs), monitoring the uptake of interventions such as expanded HIV testing and counseling, enhanced male circumcision services, and improved access to HIV care and treatment.

The dataset is available and free to download at CDC Website:https://data.cdc.gov/Global-Health/Botswana-Combination-Prevention-Project-BCPP-Publi/qcw5-4m9q/about_data.

The study is a cohort study with 3 phases. In this exercise, I only work on year 3 dataset. There are many information covered in this dataset, including demographic information of the respondent, soccioeconomic factors, HIV exposure, HIV status and conditions.

# Install all packages needed

Install and library all packages needed in this section.

```{r}


install.packages("tidyverse")
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
library(here)
install.packages("patchwork")  # This package is to redefine "/" operator for plot arrangement
library(patchwork)
install.packages("writexl")
library(writexl)
library(haven)
library(dplyr)
install.packages("janitor")
library(janitor)

```

# Loading the dataset

```{r}
## set path  dictionary 
bostwana <- here::here("cdcdata-exercise","data", "cdcbostwana.csv") # set the pathway to create relative path 

bostwana <- read_csv(bostwana) # read the dataset stored in specified path
head(bostwana)
dim(bostwana)
```
The dataset contains 11369 observations and 322 columns. 


# Data Cleaning 

I want to select interesting variables for further analysis and clean the data as needed including dropping/ lablig missing values. 

```{r}
bost_df <- bostwana %>% 
  select("region", "random_arm", "gender", "age_at_interview", "age5cat", "employment_status", "circumcised", "hiv_status_current", "hiv_status_time", "viral_load_yr3",  "cd4_survey_yr3", "cd4_surveydays_yr3", "arv_duration_days")

head(bost_df)
summary(bost_df)
```

Note: 
To make easy for further analysis, I selected several variables: 

- region = region of sbject 
- random_arm = randomized arm 
- age= age
- age5cat = age category 
- employment_status = employment status 
- hiv_status_current = Current HIV status 
- viral_load_yr3 = viral load in year 3 of the study 
- cd4_survey_yr3 = CD4 count in year 3
- cd4_surveydays_yr3 = numbers of days enrolled to survey year 3.
- arv_duration_days = number of day (duration) taking ARV 



In this excercise, I would like to see the relationship between viral load, lenght of enrollment, ARV duration, and CD4 count. Therefore, I will drop all missing data in those variables. 

```{r}

data <- bost_df %>% 
  drop_na(viral_load_yr3, cd4_survey_yr3, cd4_surveydays_yr3, arv_duration_days) %>% # I drop all observation wit N/A data
  filter(viral_load_yr3 != 0, 
         cd4_survey_yr3 != 0, 
         cd4_surveydays_yr3 != 0, 
         arv_duration_days !=0) # I found some observation contain 0 I drop those observation
```



```{r}

sapply(data, class) # I want to check class of all variables 

```


For N/A information in categorical data, I lable those missing value wih 999. However, the values are in characters. I want to change the values into factors to make us easy in analysis. 

```{r}
data <- data %>% 
  mutate(region = recode(region, 
                         "Northern" = 0,
                         "Southern" = 1, 
                         "Central" = 2), 
         random_arm = recode(random_arm, 
                             "Standard of Care" = 0,
                             "Intervention" = 1), 
         gender = recode(gender,
                         "M" = 0,
                         "F" =1), 
         age5cat = recode (age5cat, 
                           "16-24 years" = 0, 
                           "25-34 years" = 1, 
                           "35-44 years" = 2, 
                           "45-54 years" = 3, 
                           "55-64 years" = 4), 
         employment_status = recode(employment_status , 
                                    "Employed" = 0, 
                                    "Unemployed looking for work" = 1, 
                                    "Unemployed not looking for work" = 2), 
         circumcised= recode(circumcised , 
                             "No" = 0, 
                             "Yes" = 1), 
         hiv_status_current = recode(hiv_status_current, 
                                     "HIV-uninfected" = 0, 
                                     "HIV-infected" = 1, 
                                     "Refused HIV testing" = 2), 
         hiv_status_time = recode(hiv_status_time, 
                                  "HIV-negative" = 0, 
                                  "HIV-positive: previously diagnosed" = 1, 
                                  "Refused HIV testing" = 2, 
                                  "HIV-positive: newly discovered" = 3)) # this part is to recode from character to numeric to allow us convert ito factors. 

sapply(data, class) # check character of the variables 
```

Now, I want to convert categorical variables into factors

```{r}
data <- data %>%
  mutate(across (c("region", "random_arm", "gender", "age5cat", "employment_status", "circumcised", "hiv_status_current",  "hiv_status_time"), as.factor)) # this is to convert multiple variables into factors. as.factor() only can work in single variable. 

sapply(data, class) # check character of the variables 

```

I want to replace all N/A with 999 in categorical variables. 
```{r}

data <- data %>%
  mutate(across(where(is.factor), ~ replace_na(as.character(.), "999"))) %>% # replace the values
  mutate(across(where(is.character), as.factor)) # convert the variable back to factor. 

dim(data)
```

The data is clean now and ready for data exploration and analysis. There are 539 obserations and 13 variables in the final data. 

```{r}
summary_stats <- data %>%
  summarise(
    mean_viral_load = mean(viral_load_yr3, na.rm = TRUE),
    sd_viral_load = sd(viral_load_yr3, na.rm = TRUE),
    mean_cd4 = mean(cd4_survey_yr3, na.rm = TRUE),
    sd_cd4 = sd(cd4_survey_yr3, na.rm = TRUE)
  )
print(summary_stats)

```


Now I want to make summary statistics for Viral load and visualize it ina plot. 

```{r}

summary_by_arm <- data %>%
  group_by(random_arm) %>%
  summarise(
    mean_cd4 = mean(cd4_survey_yr3, na.rm = TRUE),
    sd_cd4 = sd(cd4_survey_yr3, na.rm = TRUE)
  ) # create by randomise_arm 

plot1 <- ggplot(summary_by_arm, aes(x = random_arm, y = mean_cd4, fill = random_arm)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_cd4 - sd_cd4, ymax = mean_cd4 + sd_cd4), width = 0.2) +
  labs(title = "Mean CD4 Count by randomise group with Standard Deviation",
       x = "Groups", y = "Mean CD4 Count", 
       caption = "Red (0)= Standard of Care, Green (1)= Intervention ") +
  theme_minimal() +
  theme(plot.background = element_rect(color = "black", size = 1)) 

print(plot1)


figure_file = here("cdcdata-exercise", "pictures", "CD4 Mean and SD .png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot1) # save the pictures created 
```

Now I want to make summary statistics for viral load and visualize it in a plot. 

```{r}

summary_by_arm_vl <- data %>%
  group_by(random_arm) %>%
  summarise(
    mean_vl = mean(viral_load_yr3, na.rm = TRUE),
    sd_vl = sd(viral_load_yr3, na.rm = TRUE)
  ) # create by randomise_arm 

plot2 <- ggplot(summary_by_arm_vl, aes(x = random_arm, y = mean_vl, fill = random_arm)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_vl - sd_vl, ymax = mean_vl + sd_vl), width = 0.2) +
  labs(title = "Mean Viral Load Count by randomise group with Standard Deviation",
       x = "Groups", y = "Mean Viral Load", 
       caption = "Red (0)= Standard of Care, Green (1)= Intervention ") +
  theme_minimal() +
  theme(plot.background = element_rect(color = "black", size = 1)) 

print(plot2)


figure_file = here("cdcdata-exercise", "pictures", "Viral Load Mean and SD .png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot2) # save the pictures created
```



# Data Exploration 

In this part, I want to explore the data and vizualise it before data analysis. 

```{r}
plot3 <- ggplot(data, aes(x = viral_load_yr3, y = cd4_survey_yr3, color = region)) + 
  geom_point(size = 3) +  # Specify geom as geom_point to make a scatterplot
  labs(title = "Viral Load VS CD4 count based on the region", 
       x = "Viral load", 
       y = "CD4") +  # Rename title and axes
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate state names 
        legend.position = "bottom",  # Position legend at the bottom
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold")) + # Increase size and boldness of title and axes
  scale_color_manual(values = c("0" = "#53d127", "1" = "#2ce1b0", "2" = "#ff5733"))+ # crete color manually 
  theme(plot.background = element_rect(color = "black", size = 1))

print(plot3)


figure_file = here("cdcdata-exercise", "pictures", "Viral Load VS CD4 based on region .png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot3) # save the pictures created
```



```{r}
plot4 <- ggplot(data, aes(x = arv_duration_days , y = cd4_survey_yr3, color = region)) + 
  geom_boxplot() +  # Specify geom as geom_point to make a scatterplot
  labs(title = "ARV Duration VS CD4 count based on the region", 
       x = "ARV Duration (days)", 
       y = "CD4 count",
       caption = "Green = Nothern, Blue= Northern, and red= Central") +  # Rename title and axes
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate state names 
        legend.position = "bottom",  # Position legend at the bottom
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold")) + # Increase size and boldness of title and axes
  theme(plot.background = element_rect(color = "black", size = 1)) +
  scale_x_continuous(limits = c(0, 1000)) + # Set x-axis maximum limit to 1000
  scale_color_manual(values = c("0" = "#53d127", "1" = "#2ce1b0", "2" = "#ff5733")) # crete color manually 

print(plot4)


figure_file = here("cdcdata-exercise", "pictures", "ARV Duration VS CD4 count based on the region .png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot4) # save the pictures created
```

